---
title: "Regression Models Course Project"
author: "Motor Trend"
date: "September 02, 2018"
output: pdf_document
---

## Executive Summary

In this edition of ***Motor Trend***, we look at a data set of a collection of cars from our 1974 archives. Using modern 
day data science and visualization tools, we explore the relationship between a set of variables and miles per gallon 
(MPG) and then answer the following questions for you:
1. Is an automatic or manual transmission better for MPG?
2. What is the difference in MPG for automatic and manual transmissions?

## Exploratory Analysis

The  `mtcars` data set comprises fuel consumption and 10 aspects of automobile design and performance for 32 autombiles 
(1973-74 models). Here's a glance at the first few rows of the data:

```{r preview, echo = TRUE}
data("mtcars")
head(mtcars)
```
Our key variables of interest are `am`: Transmission type (0 = automatic, 1 = manual) and `mpg`: Miles/US gallon. Let's 
take a look at the difference in means of MPG for automatic and manual transmission:

```{r mpgdiff, echo = TRUE}
aggregate(mpg ~ am, data = mtcars, FUN = mean)
```
It appears that cars with manual transmission give about 7mpg more compared to cars with automatic transmission. This 
looks like a significant difference and we will use a *t-test* to find out:

```{r t-test, echo = TRUE}
t.test(mtcars$mpg[mtcars$am == 0], mtcars$mpg[mtcars$am == 1], conf.level = 0.95)
```
Since the *p-value is less than 0.05*, we reject the null hypothesis at 95% confidence level and conclude that the MPG 
differences are indeed statistically significant.

## Regression Analysis

Let's have a closer look at the correlations of `mpg` to the other variables in `mtcars`:
```{r cor-test, echo = TRUE}
round(cor(mtcars)[, 1], 3)
```
From the correlation data, there is a strong positive and inverse correlation between MPG and the other variables. Next, 
we will take a look at the regression model for MPG and transmission. 

```{r reg-model, echo = TRUE}
fit1 <- lm(mpg ~ am, data = mtcars)
summary(fit1)
```
This regression model shows that the average MPG for automatic transmission is 17.1, and the MPG increases by 7.2 for 
manual transmission. However, the *R^2^* value is 0.36, which indicates this model only explains 36% of the variance. 
Thus we build a few multivariate linear regression models. 

```{r reg-model-add, echo = TRUE}
fit2 <- lm(mpg ~ am + cyl, data = mtcars)
fit3 <- lm(mpg ~ am + cyl + disp, data = mtcars)
fit4 <- lm(mpg ~ am + cyl + disp + hp, data = mtcars)
fit5 <- lm(mpg ~ am + cyl + disp + hp + wt, data = mtcars)
fit6 <- lm(mpg ~ am + cyl + disp + hp + wt + drat, data = mtcars)
fit7 <- lm(mpg ~ am + cyl + disp + hp + wt + drat + vs, data = mtcars)
anova(fit1, fit2, fit3, fit4, fit5, fit6, fit7)
```
From the probabilities in the above anova table, we can say that `am`, `cyl`, `hp` and `wt` are the significant 
predictors of `mpg` at 95% confidence interval. 

```{r better-model, echo = TRUE}
better_model <- lm(mpg ~ am + cyl + hp + wt, data = mtcars)
summary(better_model)
```

## Conclusions

1. The above multivariate model shows that cars with manual transmission give better MPG compared to automatic 
transmission. 
2. In contrast to our initial regression model which shows a huge difference in mileage for manual and automatic cars, 
the multivariate model indicates that cars with manual transmission give 1.48mpg higher than automatic transmission cars 
when other variables like number of cylinders (`cyl`), gross horsepower (`hp`) and weight (`wt`) are also taken into 
account. Moreover, this is a better regression model as it explains 85% of the variance. 

## Appendix

#### Plot - 1: Boxplot of MPG by transmission type
```{r boxplot, echo = FALSE, fig.align = "center", fig.height = 5, fig.width = 8}
boxplot(mpg ~ am, data = mtcars, col = c("red", "blue"), xlab = "Transmission type", 
        ylab = "Miles per gallon (US)")
```

#### Plot - 2: Correlation heatmap of all variables
```{r heatmap, echo = FALSE, fig.align = "center", fig.height = 5, fig.width = 8}
suppressMessages(library(gplots))
suppressMessages(library(RColorBrewer))
data.plot <- round(cor(mtcars), 3)
col_palette <- colorRampPalette(c("blue", "white", "red"))(n = 299)
col_breaks <- c(seq(-1, 0, length = 100), seq(0.01, 0.8, length = 100), seq(0.81, 1 , length = 100))
heatmap.2(data.plot, cellnote = data.plot, margins = c(4, 0), notecol = "black", density.info = "none", 
          trace = "none", col = col_palette, breaks = col_breaks, dendrogram = "none", keysize = 1, 
          lmat = rbind(c(6, 1, 2), c(5, 4, 3)), lwid = c(1, 10, 1), lhei = c(5, 1.5))
```

#### Plot - 3: Analysis of residuals
```{r analysis, echo = FALSE, fig.align = "center", fig.height = 5, fig.width = 8}
par(mfrow = c(2, 2))
plot(better_model)
```