---
title: "An analysis of severe weather events on U.S. health and economy"
date: "November 16, 2017"
output: html_document
---

------------------------------------------------------------------------------------------------------------------

# Synopsis 

This report explores and analyzes data on severe weather events obtained from the U.S. National Oceanic and Atmospheric
Administration's (NOAA) Storm database. The data can be downloaded from [this link][1]. The events in the database 
start in the year 1950 and end in November 2011. Of particular interest to us in this analysis are observing the damage 
to life and economy caused by different weather events. The most catastrophic weather events with respect to public 
health and economic impact are then identified. This would aid in prioritizing resources for different types of events. 

# Data Processing

Contents of the data were viewed and analyzed with the help of a software package called **RStudio**. The R code 
downloads the zip file containing the data for you, extracts the contents of the file and loads the data into a 
readable tabular format using the **data.table** package. The code to do this is shown in the code chunk below:

```{r Load, echo = TRUE, cache = TRUE, warning = FALSE, message = FALSE}
rm(list = ls())         # remove any variables in global environment 
library(data.table)     # load data.table package

url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"   # set download link to the database
download.file(url, destfile = "./StormData.csv.bz2")            # download zip file with name as "StormData"

data.in <- data.table(read.csv("StormData.csv.bz2", stringsAsFactors = FALSE))      # read data into R
```

As far as the scope of this report goes, not all columns in the raw dataset are required. The raw dataset is broken 
down into 2 subsets: one to assess the impact of weather events on public health and another to assess the economic 
impact. 

The code chunk dealing with the public health subset, named "harm.health" is displayed below: 

```{r Health, echo = TRUE, warning = FALSE, message = FALSE}
# create subset 1 by selecting event type, fatalities and injuries from raw dataset:
harm.health <- subset(data.in, select = c("EVTYPE", "FATALITIES", "INJURIES"))

# create a new column equal to the sum of fatalities and injuries:
harm.health$"TOTAL" <- harm.health$FATALITIES + harm.health$INJURIES

# calculate the sum of fatalities, injuries and total number of affected people for each event type:
harm.health <- aggregate(cbind(FATALITIES, INJURIES, TOTAL) ~ EVTYPE, harm.health, sum)

# arrange subset 1 in decreasing order of total number of affected people:
harm.health <- harm.health[order(-harm.health$TOTAL), ]
```

For the economic impact assessment, the damages are expressed in character values denoting hundreds, thousands, 
millions and billions. For a significant number of observations, the damages are not well documented or some garbage 
value is assigned. 

The code chunk for appropriately converting the character values to numeric values and obtaining subset, named 
"econ.clean", that deals with economic damage is displayed below:

```{r Economy, echo = TRUE, warning = FALSE, message = FALSE}
# take subset of event type, property and crop damage, property damage and crop damage multiplier exponent:
econ.data <- subset(data.in, select = c("EVTYPE", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP"))

# create new column containing appropriate numeric values as depicted by character values of property damage exponent: 
econ.data$PMULT[econ.data$PROPDMGEXP == "" | econ.data$PROPDMGEXP == "-" | econ.data$PROPDMGEXP == "?" | 
                          econ.data$PROPDMGEXP == "+"] <- 0
econ.data$PMULT[econ.data$PROPDMGEXP == "0"] <- 10^0
econ.data$PMULT[econ.data$PROPDMGEXP == "1"] <- 10^1
econ.data$PMULT[econ.data$PROPDMGEXP == "2" | econ.data$PROPDMGEXP == "h" | econ.data$PROPDMGEXP == "H"] <- 10^2
econ.data$PMULT[econ.data$PROPDMGEXP == "3" | econ.data$PROPDMGEXP == "k" | econ.data$PROPDMGEXP == "K"] <- 10^3
econ.data$PMULT[econ.data$PROPDMGEXP == "4"] <- 10^4
econ.data$PMULT[econ.data$PROPDMGEXP == "5"] <- 10^5
econ.data$PMULT[econ.data$PROPDMGEXP == "6" | econ.data$PROPDMGEXP == "m" | econ.data$PROPDMGEXP == "M"] <- 10^6
econ.data$PMULT[econ.data$PROPDMGEXP == "7"] <- 10^7
econ.data$PMULT[econ.data$PROPDMGEXP == "8"] <- 10^8
econ.data$PMULT[econ.data$PROPDMGEXP == "b" | econ.data$PROPDMGEXP == "B"] <- 10^9

# create new column containing appropriate numeric values as depicted by character values of crop damage exponent:
econ.data$CMULT[econ.data$CROPDMGEXP == "" | econ.data$CROPDMGEXP == "?"] <- 0
econ.data$CMULT[econ.data$CROPDMGEXP == "0"] <- 10^0
econ.data$CMULT[econ.data$CROPDMGEXP == "2"] <- 10^2
econ.data$CMULT[econ.data$CROPDMGEXP == "k" | econ.data$CROPDMGEXP == "K"] <- 10^3
econ.data$CMULT[econ.data$CROPDMGEXP == "m" | econ.data$CROPDMGEXP == "M"] <- 10^6
econ.data$CMULT[econ.data$CROPDMGEXP == "B"] <- 10^9

# calculate total amount of damage to property and damage to crops:
econ.data$PROPERTY <- econ.data$PROPDMG * econ.data$PMULT
econ.data$CROP <- econ.data$CROPDMG * econ.data$CMULT

# create subset 2 by selecting event type, property damage amount and crop damage amount from "econ.data" subset:
econ.clean <- subset(econ.data, select = c("EVTYPE", "PROPERTY", "CROP"))

# create a new column equal to the sum of property damage amount and crop damage amount:
econ.clean$"TOTAL" <- econ.clean$PROPERTY + econ.clean$CROP

# calculate the sum of property, crop and overall damage amounts for each event type:
econ.clean <- aggregate(cbind(PROPERTY, CROP, TOTAL) ~ EVTYPE, econ.clean, sum)

# arrange subset 2 in decreasing order of total property and crop damage amount:
econ.clean <- econ.clean[order(-econ.clean$TOTAL), ]
```

# Results

A snapshot of the top 10 weather events that have the most severe impact on population health and economic damage is 
shown below:

```{r Tables, echo = TRUE, warning = FALSE, message = FALSE}
head(harm.health, n = 10)
head(econ.clean, n = 10)
```

```{r Maximums, echo = TRUE, warning = FALSE, message = TRUE}
# event causing most number of fatalities 
harm.health[which.max(harm.health$FATALITIES), ]

# event causing most number of injuries 
harm.health[which.max(harm.health$INJURIES), ]

# event causing most number of fatalities and injuries combined
harm.health[which.max(harm.health$TOTAL), ]

# event causing most damage to property
econ.clean[which.max(econ.clean$PROPERTY), ]

# event causing most damage to crops
econ.clean[which.max(econ.clean$CROP), ]

# event causing most damage to property and crops combined
econ.clean[which.max(econ.clean$TOTAL), ]
```

From these tables, we can observe that:  
- **Tornadoes** cause the most number of fatalities  
- **Tornadoes** cause the most number of injuries  
- **Tornadoes** cause the most number of fatalities combined  
- **Floods** cause the most damage to property  
- **Droughts** cause the most damage to crops  
- **Floods** cause the most damage to property and crops combined  

Let us take a look at the names of all severe weather events:

```{r Names, echo = TRUE, warning = FALSE, message = FALSE}
harm.health$EVTYPE
```

If you take a closer look at the names of the weather events, you will observe that names of many events are repeated 
(lower or upper case) or look similar. Some of them even are even spelt incorrectly.  

Next, an attempt is made to choose the top 6 severe events (excluding events with similar names), search for events 
with names similar to these 6 events and group all the into events into events with a common name.  

For the events affecting population health, these 6 events are chosen: **TORNADO, HEAT, WIND, FLOODING, LIGHTNING, 
STORM**  
For the events affecting economy, these 6 events are chosen: **FLOODING, HURRICANE, TORNADO, STORM, HAIL, DROUGHT**  

We must note that some of these events will overlap and will be repeated, but it will be interesting to see if the 
order changes when the events are grouped under a common name.  

```{r PopGroup, echo = TRUE, warning = FALSE, message = FALSE}
# create an empty data table, data1, to store events affecting population health
data1 <- data.frame()

# search for all events containing the word "torn" and classify them under "TORNADO"
data.temp <- harm.health[grep("torn", harm.health$EVTYPE, ignore.case = TRUE), ]
data.temp$EVTYPE <- "TORNADO"
data1 <- rbind(data1, data.temp)

# search for all events containing the word "heat" and classify them under "HEAT"
data.temp <- harm.health[grep("heat", harm.health$EVTYPE, ignore.case = TRUE), ]
data.temp$EVTYPE <- "HEAT"
data1 <- rbind(data1, data.temp)

# search for all events containing the word "wind" and classify them under "WIND"
data.temp <- harm.health[grep("wind", harm.health$EVTYPE, ignore.case = TRUE), ]
data.temp$EVTYPE <- "WIND"
data1 <- rbind(data1, data.temp)

# search for all events containing the word "flood" and classify them under "FLOODING"
data.temp <- harm.health[grep("flood", harm.health$EVTYPE, ignore.case = TRUE), ]
data.temp$EVTYPE <- "FLOODING" 
data1 <- rbind(data1, data.temp)

# search for all events containing the word "lightn" and classify them under "LIGHTNING"
data.temp <- harm.health[grep("lightn", harm.health$EVTYPE, ignore.case = TRUE), ]
data.temp$EVTYPE <- "LIGHTNING"
data1 <- rbind(data1, data.temp)

# search for all events containing the word "storm" and classify them under "STORM"
data.temp <- harm.health[grep("storm", harm.health$EVTYPE, ignore.case = TRUE), ]
data.temp$EVTYPE <- "STORM"
data1 <- rbind(data1, data.temp)

# calculate the sum of fatalities, injuries and total number of affected people for each event type:
data1 <- aggregate(cbind(FATALITIES, INJURIES, TOTAL) ~ EVTYPE, data1, sum)

# arrange data 1 in decreasing order of total number of affected people:
data1 <- data1[order(-data1$TOTAL), ]
```

```{r EconGroup, echo = TRUE, warning = FALSE, message = FALSE}
# create an empty data table, data2, to store events affecting economy
data2 <- data.frame()

# search for all events containing the word "flood" and classify them under "FLOODING"
data.temp <- econ.clean[grep("flood", econ.clean$EVTYPE, ignore.case = TRUE), ]
data.temp$EVTYPE <- "FLOODING"
data2 <- rbind(data2, data.temp)

# search for all events containing the word "hurr" and classify them under "HURRICANE"
data.temp <- econ.clean[grep("hurr", econ.clean$EVTYPE, ignore.case = TRUE), ]
data.temp$EVTYPE <- "HURRICANE"
data2 <- rbind(data2, data.temp)

# search for all events containing the word "torn" and classify them under "TORNADO"
data.temp <- econ.clean[grep("torn", econ.clean$EVTYPE, ignore.case = TRUE), ]
data.temp$EVTYPE <- "TORNADO" 
data2 <- rbind(data2, data.temp)

# search for all events containing the word "storm" and classify them under "STORM"
data.temp <- econ.clean[grep("storm", econ.clean$EVTYPE, ignore.case = TRUE), ]
data.temp$EVTYPE <- "STORM"
data2 <- rbind(data2, data.temp)

# search for all events containing the word "hail" and classify them under "HAIL"
data.temp <- econ.clean[grep("hail", econ.clean$EVTYPE, ignore.case = TRUE), ]
data.temp$EVTYPE <- "HAIL"
data2 <- rbind(data2, data.temp)

# search for all events containing the word "droug" and classify them under "DROUGHT"
data.temp <- econ.clean[grep("droug", econ.clean$EVTYPE, ignore.case = TRUE), ]
data.temp$EVTYPE <- "DROUGHT"
data2 <- rbind(data2, data.temp)

# calculate the sum of property, crop and overall damage amounts for each event type:
data2 <- aggregate(cbind(PROPERTY, CROP, TOTAL) ~ EVTYPE, data2, sum)

# arrange subset 2 in decreasing order of total property and crop damage amount:
data2 <- data2[order(-data2$TOTAL), ]
```

A snapshot of the top 6 weather events after grouping is shown below:

```{r Grouped, echo = TRUE, warning = FALSE, message = FALSE}
data1
data2
```

We see that the order has slightly changed, but the most severe events have still remained at the top.  

The following multiple bar plots provide a pictorial visualization of the grouped data:

```{r Plot1, echo = TRUE, warning = FALSE, message = FALSE, fig.height = 14, fig.width = 20, fig.cap = "Figure 1"}
library(data.table)     # load data.table package
library(ggplot2)        # load ggplot2 package
# make bar plot for top 6 severe events affecting population health
data1 <- melt(data1, id.vars = "EVTYPE")
print(ggplot(data1, aes(x = EVTYPE, y = value)) + geom_bar(aes(fill = variable), width = 0.4, 
                                                           position = position_dodge(width = 0.5), stat = "identity") + 
          ggtitle("Top 6 catastrophic events with respect to population health") + 
          theme(plot.title = element_text(size = 24, face = "bold")) + 
          theme(axis.text.x = element_text(size = 18, hjust = 1, vjust = 1, angle = 45), 
                axis.title.x = element_text(size = 22, face = "bold")) + 
          theme(axis.text.y = element_text(size = 18), axis.title.y = element_text(size = 22, face = "bold")) +
          theme(legend.text = element_text(size = 18), legend.key.height = unit(3, "lines"), 
                legend.title = element_text(size = 22, face = "bold")) + 
          ylab("Count") + xlab("Event Type"))
```

```{r Plot2, echo = TRUE, warning = FALSE, message = FALSE, fig.height = 14, fig.width = 20, fig.cap = "Figure 2"}
library(data.table)     # load data.table package
library(ggplot2)        # load ggplot2 package
# make bar plot for top 6 severe events causing economic loss
data2 <- melt(data2, id.vars = "EVTYPE")
data2$value <- data2$value/10^9
print(ggplot(data2, aes(x = EVTYPE, y = value)) + geom_bar(aes(fill = variable), width = 0.4, 
                                                           position = position_dodge(width = 0.5), stat = "identity") + 
          ggtitle("Top 6 catastrophic events in terms of economic damage") + 
          theme(plot.title = element_text(size = 24, face = "bold")) + 
          theme(axis.text.x = element_text(size = 18, hjust = 1, vjust = 1, angle = 45), 
                axis.title.x = element_text(size = 22, face = "bold")) + 
          theme(axis.text.y = element_text(size = 18), axis.title.y = element_text(size = 22, face = "bold")) +
          theme(legend.text = element_text(size = 18), legend.key.height = unit(3, "lines"), 
                legend.title = element_text(size = 22, face = "bold")) + 
          ylab("Loss ( in billion USD )") + xlab("Event Type"))
```

A published report on RPubs can be found at [this link][2].

<!--Set links below-->

[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2  

[2]: http://rpubs.com/octopacks/WeatherReport 